---
output:
  html_document:
    keep_md: yes
---
## Reproducible Research: Peer Assessment 2

## U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database analysis

## SYNOPSIS

This report consists in analyzing the NOAA storm database containing data on extreme climate events. This data was collected during the period from 1950 through 2011. The purpose of this analysis is to answer the following two questions:

- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

- Across the United States, which types of events have the greatest economic consequences?


## Data Processing

```{r}
setwd( "C:/Users/soujanya/Desktop/Coursera")
install.packages("R.utils",repos="http://cran.rstudio.com/")
library(R.utils)
# load data into R
storm <- read.csv("repdata-data-StormData.csv")
head(storm)
```

### subset the data to health and economic impact analysis against weather event
```{r}
Event <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", 
    "CROPDMGEXP")
required_data <- storm[Event]
```

### Preparing the property damage data

#### exploring the property exponent
```{r}
unique(required_data$PROPDMGEXP)
```

### Sorting the property exponent data
```{r}
required_data$PROPEXP[required_data$PROPDMGEXP == "K"] <- 1000
required_data$PROPEXP[required_data$PROPDMGEXP == "M"] <- 1e+06
required_data$PROPEXP[required_data$PROPDMGEXP == ""] <- 1
required_data$PROPEXP[required_data$PROPDMGEXP == "B"] <- 1e+09
required_data$PROPEXP[required_data$PROPDMGEXP == "m"] <- 1e+06
required_data$PROPEXP[required_data$PROPDMGEXP == "0"] <- 1
required_data$PROPEXP[required_data$PROPDMGEXP == "5"] <- 1e+05
required_data$PROPEXP[required_data$PROPDMGEXP == "6"] <- 1e+06
required_data$PROPEXP[required_data$PROPDMGEXP == "4"] <- 10000
required_data$PROPEXP[required_data$PROPDMGEXP == "2"] <- 100
required_data$PROPEXP[required_data$PROPDMGEXP == "3"] <- 1000
required_data$PROPEXP[required_data$PROPDMGEXP == "h"] <- 100
required_data$PROPEXP[required_data$PROPDMGEXP == "7"] <- 1e+07
required_data$PROPEXP[required_data$PROPDMGEXP == "H"] <- 100
required_data$PROPEXP[required_data$PROPDMGEXP == "1"] <- 10
required_data$PROPEXP[required_data$PROPDMGEXP == "8"] <- 1e+08
#### give 0 to invalid exponent data, so they not count in
required_data$PROPEXP[required_data$PROPDMGEXP == "+"] <- 0
required_data$PROPEXP[required_data$PROPDMGEXP == "-"] <- 0
required_data$PROPEXP[required_data$PROPDMGEXP == "?"] <- 0
#### compute the property damage value
required_data$PROPDMGVAL <- required_data$PROPDMG * required_data$PROPEXP
```

### Preparing the crop damage data

#### exploring the crop exponent data
```{r}
unique(required_data$CROPDMGEXP)
```

### Sorting the property exponent data
```{r}
required_data$CROPEXP[required_data$CROPDMGEXP == "M"] <- 1e+06
required_data$CROPEXP[required_data$CROPDMGEXP == "K"] <- 1000
required_data$CROPEXP[required_data$CROPDMGEXP == "m"] <- 1e+06
required_data$CROPEXP[required_data$CROPDMGEXP == "B"] <- 1e+09
required_data$CROPEXP[required_data$CROPDMGEXP == "0"] <- 1
required_data$CROPEXP[required_data$CROPDMGEXP == "k"] <- 1000
required_data$CROPEXP[required_data$CROPDMGEXP == "2"] <- 100
required_data$CROPEXP[required_data$CROPDMGEXP == ""] <- 1
#### give 0 to invalid exponent data, so they not count in
required_data$CROPEXP[required_data$CROPDMGEXP == "?"] <- 0
#### compute the crop damage value
required_data$CROPDMGVAL <- required_data$CROPDMG * required_data$CROPEXP
```

### Aggregate the data by event

#### aggregate the data by event
```{r}
fatal <- aggregate(FATALITIES ~ EVTYPE, data = required_data, FUN = sum)
injury <- aggregate(INJURIES ~ EVTYPE, data = required_data, FUN = sum)
propdmg <- aggregate(PROPDMGVAL ~ EVTYPE, data = required_data, FUN = sum)
cropdmg <- aggregate(CROPDMGVAL ~ EVTYPE, data = required_data, FUN = sum)
```

## Results 
### Aross the United States, Which types of events are nost harmful with respect to population health?

#### get top10 event with highest fatalities
```{r}
fatal10 <- fatal[order(-fatal$FATALITIES), ][1:10, ]
#### get top10 event with highest injuries
injury10 <- injury[order(-injury$INJURIES), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(fatal10$FATALITIES, las = 3, names.arg = fatal10$EVTYPE, main = "Weather Events With The Top 10 Highest Fatalities", 
    ylab = "number of fatalities", col = "green")
barplot(injury10$INJURIES, las = 3, names.arg = injury10$EVTYPE, main = "Weather Events With the Top 10 Highest Injuries", 
    ylab = "number of injuries", col = "green")
```

#### The most harmful weather event to population health is Tornado.It is cause for both the highest fatalities and the highest injuries across United States.

### Across the United States, which types of events have the greatest economic consequences?

#### get top 10 events with highest property damage
```{r}
propdmg10 <- propdmg[order(-propdmg$PROPDMGVAL), ][1:10, ]
#### get top 10 events with highest crop damage
cropdmg10 <- cropdmg[order(-cropdmg$CROPDMGVAL), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(propdmg10$PROPDMGVAL/(10^9), las = 3, names.arg = propdmg10$EVTYPE, 
    main = "Top 10 Events with Greatest Property Damages", ylab = "Cost of damages ($ billions)", 
    col = "green")
barplot(cropdmg10$CROPDMGVAL/(10^9), las = 3, names.arg = cropdmg10$EVTYPE, 
    main = "Top 10 Events With Greatest Crop Damages", ylab = "Cost of damages ($ billions)", 
    col = "green")
```

#### The weather events have the greatest economic consequences are: flood, drought, Tornado and Typhoon.

#### Across the United States, flood, tornado and typhoon have caused the greatest damage to properties. Drought and flood come as the causes for the greatest damage to crops.